The effect of climate forcing on numerical simulations of the Cordilleran Ice Sheet at the Last Glacial Maximum

### 2013-07-08 sections/intro.tex
Apply Irina's corrections in the intro

Arguably the most physically sound way to force an ice sheet model in the past is to couple it with a General Circulation Model (GCM) \citep{yoshimori-etal-2001,calov-etal-2002,abeouchi-etal-2007,charbit-etal-2013}. However the computational demand of GCMs is such that only models of intermediate complexity can run on the time-scales of tens of thousands of years characteristic of ice sheet growth and decay.
\julien{Add words on limitations of GCM in this context?}

Climatologies obtained from uncoupled GCM palaeo-climate simulations such as produced within the PMIP project \citep{joussaume-taylor-1995} provide a more accurate representation of past climate and are commonly used to force ice sheet models. Because they are only available for specific periods of time, they require either an assumption of steady-state \citep{huybrechts-tsiobbel-1996}, or interpolation through time between climatologies from different periods, which can be linear \citep{charbit-etal-2002}, or modulated by a ``glacial index'' weighting function derived from ice core $\delta^{18}$\,O records \citep{marshall-clarke-1999,tarasov-peltier-2004,zweck-huybrechts-2005,gregoire-etal-2012}. An important drawback in this approach is that palaeo-climate simulations themselves rely on global ice sheet reconstructions such as the ICE-4G \citep{peltier-1994} for their surface topographic boundary condition. This indirectly controls subsequently modelled ice sheet geometries, whose spatial extent tend to conform to these reconstructions.
\julien[noline]{Should I include references not covering the Cordilleran Ice Sheet?}

@@

Here, we propose to address some of the uncertainties concerning climate forcing of numerical glacier models by evaluating their response, in terms of glacier extent, to inputs from several climate datasets. Few studies of this kind are presently available. \citet{quiquet-etal-2012} assessed the sensitivity of a Greenland ice sheet model to various atmospheric forcing, including a regional parametrization \citep{fausto-etal-2009}, output from several GCMs and a climate reanalysis. \citet{rodgers-etal-2004,charbit-etal-2007} tested the sensitivity of a model of the Northern Hemisphere ice sheets to different PMIP LGM simulations. Here, to limit degrees of freedom in our model, and obtain results independent from palaeo-ice sheet reconstructions such as the ICE-4G, we use a simple temperature offset approach similar to \citep{greve-etal-1999,bintanja-etal-2005} and assess ice sheet model sensitivity to the choice of present-day climate data. Rather than using GCM output, we force our model with climate reanalysis data, which include observational information through data assimilation \citep{bengtsson-etal-2007}. Furthermore, we focus our study regionally to the former Cordilleran Ice Sheet in Northern America.

The Cordilleran Ice Sheet (Fig.~\ref{fig:locmap}) covered an area that presently experiences strong regional variations in climate. In a numerical modelling perspective, it is one of the least studied palaeo-ice sheets of the Northern Hemisphere, despite the fact that significant geomorphological data is available to constrain its extent \citep{jackson-clague-1991,dukrodkin-1999,kaufman-manley-2004,kleman-etal-2010,margold-etal-2011}. Although the Cordilleran Ice Sheet was not the primary target of these simulations, it has previously been modelled as part of efforts to model ice sheets in North America \citep{marshall-clarke-1999,calov-etal-2002,tarasov-peltier-1997,tarasov-peltier-2004,gregoire-etal-2012}, the Northern Hemisphere \citep{huybrechts-tsiobbel-1996,greve-etal-1999,charbit-etal-2002,charbit-etal-2007,charbit-etal-2013,johnson-fastook-2002,rodgers-etal-2004,bintanja-etal-2005,zweck-huybrechts-2005,abeouchi-etal-2007} or world-wide \citep{yoshimori-etal-2001}. While these studies reproduce the magnitude of North American glaciation at LGM reasonably well, there exist a tendency in simulations independent from ice sheet reconstructions such as the ICE-4G to predict excessive ice cover in parts of northern Yukon Territory and continental Alaska that have remained ice-free throughout the Pleistocene \citep{dukrodkin-1999,kaufman-manley-2004}.

Here we use a PISM, a Parallel Ice Sheet Model \citep{web:pism} \julien{Citation recommended by \url{http://www.pism-docs.org/wiki/doku.php?id=citing_pism}} to simulate the extent and thickness of Cordilleran Ice Sheet at the LGM. We force our model with multiple climate datasets and compare our results to the mapped LGM ice sheet margin from \citet{dyke-2004}. To our knowledge, this is the first modelling study which specifically focus on the Cordilleran Ice Sheet since the one by \citet{robert-1991}. While the modelling domain of \citet{robert-1991} covered only the southern part of the ice sheet, our model also includes higher resolution and an improved treatment of ice thermodynamics and bedrock response due to large advances made in ice sheet modelling since then. Model set-up is presented in section~\ref{sec:model} and climate forcing in section~\ref{sec:climate}. Results are exposed in section~\ref{sec:results} and discussed in section~\ref{sec:discussion}.

>>>>

Arguably the most physically sound way to force an ice sheet model in the past is to couple it with a General Circulation Model (GCM) \citep{yoshimori-etal-2001,calov-etal-2002,abeouchi-etal-2007,charbit-etal-2013}. However the computational demand of GCMs is such that only models of intermediate complexity can run on the time-scales of tens of thousands of years characteristic of ice sheet growth and decay.

Climatologies obtained from uncoupled GCM palaeo-climate simulations such as produced within the PMIP project \citep{joussaume-taylor-1995} provide a more accurate representation of past climate and are commonly used to force ice sheet models. Because they are only available for specific periods of time, they require either an assumption of steady-state \citep{huybrechts-tsiobbel-1996}, or interpolation through time between climatologies from different periods, which can be linear \citep{charbit-etal-2002}, or modulated by a ``glacial index'' weighting function derived from ice core $\delta^{18}$\,O records \citep{marshall-clarke-1999,tarasov-peltier-2004,zweck-huybrechts-2005,gregoire-etal-2012}. An important drawback in this approach is that palaeo-climate simulations themselves rely on global ice sheet reconstructions such as the ICE-4G \citep{peltier-1994} for their surface topographic boundary condition. This exerts indirect influence on subsequently modelled ice sheet geometries, whose spatial extent tends to conform to these reconstructions.
\julien[noline]{Should I include references not covering the Cordilleran Ice Sheet?}

@@

Here, we propose to address some of the uncertainties concerning climate forcing of numerical glacier models by evaluating their responses, in terms of glacier extent, to inputs from several climate datasets. Few studies of this kind are presently available. \citet{quiquet-etal-2012} assessed the sensitivity of a Greenland ice sheet model to various atmospheric forcing, including a regional parametrization \citep{fausto-etal-2009}, output from several GCMs and a climate reanalysis. \citet{rodgers-etal-2004,charbit-etal-2007} tested the sensitivity of a model of the Northern Hemisphere ice sheets to different PMIP LGM simulations. Here, to limit degrees of freedom in our model, and obtain results independent of palaeo-ice sheet reconstructions such as the ICE-4G, we use a simple temperature offset approach similar to \citep{greve-etal-1999,bintanja-etal-2005} and assess ice sheet model sensitivity to the choice of present-day climate data. Rather than using GCM output, we force our model with climate reanalysis data, which include observational information through data assimilation \citep{bengtsson-etal-2007}. Furthermore, we focus our study regionally on the former Cordilleran Ice Sheet in Northern America.

The Cordilleran Ice Sheet (Fig.~\ref{fig:locmap}) covered an area that presently experiences strong regional variations in climate. In a numerical modelling perspective, it is one of the least studied palaeo-ice sheets of the Northern Hemisphere, despite the fact that significant geomorphological data is available to constrain its extent \citep{jackson-clague-1991,dukrodkin-1999,kaufman-manley-2004,kleman-etal-2010,margold-etal-2011}. Although the Cordilleran Ice Sheet was not the primary target of these simulations, it has previously been modelled as part of efforts to model ice sheets in North America \citep{marshall-clarke-1999,calov-etal-2002,tarasov-peltier-1997,tarasov-peltier-2004,gregoire-etal-2012}, the Northern Hemisphere \citep{huybrechts-tsiobbel-1996,greve-etal-1999,charbit-etal-2002,charbit-etal-2007,charbit-etal-2013,johnson-fastook-2002,rodgers-etal-2004,bintanja-etal-2005,zweck-huybrechts-2005,abeouchi-etal-2007} or world-wide \citep{yoshimori-etal-2001}. While these studies reproduce the magnitude of North American glaciation at LGM reasonably well, there exists a tendency in the simulations that are independent of ice sheet reconstructions such as the ICE-4G to predict excessive ice cover in parts of northern Yukon Territory and continental Alaska that have remained ice-free throughout the Pleistocene \citep{dukrodkin-1999,kaufman-manley-2004}.

Here we use a PISM, a Parallel Ice Sheet Model \citep{web:pism} \julien{Citation recommended by \url{http://www.pism-docs.org/wiki/doku.php?id=citing_pism}}, to simulate the extent and thickness of Cordilleran Ice Sheet at the LGM. We force our model with multiple climate datasets and compare our results to the mapped LGM ice sheet margin from \citet{dyke-2004}. To our knowledge, this is the first modelling study that specifically focuses on the Cordilleran Ice Sheet since the one by \citet{robert-1991}. While the modelling domain of \citet{robert-1991} covered only the southern part of the ice sheet, our model also includes higher resolution and an improved treatment of ice thermodynamics and bedrock response due to large advances made in ice sheet modelling since then. Model set-up is presented in section~\ref{sec:model} and climate forcing in section~\ref{sec:climate}. Results are exposed in section~\ref{sec:results} and discussed in section~\ref{sec:discussion}.
